Abstractions for Multi-Sorted Substitutions

### 2024-03-11 paper/paper.tex
Update.

  % Before we define \ACode{\ACon{ComposeKit}} and
  % \ACode{\ACon{ComposeTraversal}}, we need one more ingredient.

@@

  % - Reason about multiple applications in terms of a single application
  % - e.g. wk-cancels-⦅⦆ or dist-↑-⦅⦆, often important for induction hypothesis
  % - Mcbride uses tactics
  % - We introduce an abstraction similar to kits
  % - ComposeKit has primitives for composition, ComposeTraversal has the property

@@

  To fill this gap, we introduce a new abstraction that we call a
  \ACode{\ACon{ComposeKit}},
  which provides an operation \ACode{\AField{\_\&/⋯\_}} that unifies
  \ACode{\ADef{\_\&\_}} and \ACode{\ADef{\_⋯\_}}. This allows us to
  define a general composition, which informally can be described as:
  \ExampleCode\KComposition

  % \ACode{\ACon{ComposeKit}} and \ACode{\ADef{\_·ₖ\_}} are analogous to
  % \ACode{\ACon{Kit}} and \ACode{\ADef{\_→ₖ\_}}, whereas
  % \ACode{\ACon{ComposeTraversal}} and \ACode{\AField{⋯-fusion}} are analogous to
  % \ACode{\ACon{ComposeTraversal}} and \ACode{\AField{\_⋯\_}}.

@@

  In this subsection, we define the \ACode{\ACon{ComposeKit}}
  abstraction, which provides the \ACode{\AField{\_\&/⋯\_}} operation
  that is necessary for defining composition, and the
  \ACode{\ACon{ComposeTraversal}}, which provides the
  \ACode{\AField{⋯-fusion}} lemma for general maps.

  Similarly as it was the case for \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}},
  the lemma of the \ACode{\ACon{ComposeTraversal}} needs to be proved for
  an abstract \ACode{\ACon{ComposeKit}} and in return defines the
  four concrete \ACode{\ACon{ComposeKit}} instances for us. This allows us to break
  the dependencies between the different \ACode{\AField{⋯-fusion}} lemmas,
  similar to how \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}} allowed
  us to break the dependency between traversals with renamings and substitutions.

  For this purpose, the \ACode{\ACon{ComposeKit}} abstraction, needs
  to satisfy two requirements:
  it needs to provide us with enough machinery to define composition, but
  it also needs to provide us with enough axioms that we are able to instanciate
  the \ACode{\ACon{ComposeTraversal}}. It is defined as followed:
  \LibCode\KComposeKit
  The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
  (according to \ACode{K₁}) and a renaming or substitution
  (according to \ACode{K₂}) and performs the map traversal returning
  a variable or term (according to \ACode{K₁⊔K₂}).
  The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency and is completely determined by the choice of \ACode{K₁}
  and \ACode{K₂}.

@@

  Similar as for the \ACode{\ACon{Kit}}, we derive further definitions
  from the axioms of a \ACode{\ACon{ComposeKit}}:
  \begin{itemize}
  \item
    We define map composition as motivated by the previous subsection:
    \LibCode\KComposition
  \item
    We show that the \ACode{\AField{\_\&/⋯\_}} operator behaves like
    \ACode{\AField{\_\&\_}} when applied to variables:
    \LibCode\KComposeKitAp
    % \LibCode\KComposeKitApProof

    % The \ACode{\AField{`/id}} on the left side is from \ACode{K₁⊔K₂}, whereas
    % the \ACode{\AField{`/id}} on the right side is from \ACode{K₂}.
  \item
    We show that lifting \ACode{\AField{\_↑\_}} distributes over
    composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}
  To make it easier to explicitly specify a \ACode{\ACon{ComposeKit}},
  we define the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}
  to stand for \ACode{ϕ₁ \ADef{·ₖ} ϕ₂} for some
  \ACode{\ACon{ComposeKit}} \ACode{C}.
  % \LibCode\KComposeKitNotation

  We then define the \ACode{\ACon{ComposeTraversal}} structure,
  which provides the \ACode{\AField{⋯-fusion}} lemma for arbitrary
  \ACode{\ACon{ComposeKit}}s:
  \LibCode\KComposeTraversal

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the axiom of a \ACode{\ACon{ComposeTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    We define \ACode{\ACon{ComposeKit}} instances for each of the four
    combinations between renamings and substitutions:
    \LibCode\KComposeKitInstances
    Interestingly, the \ACode{\ADef{Cₛ}} instance has another
    \ACode{\ADef{ComposeKit}} parameter.
    This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
    which yields the \ACode{\ADef{ComposeKit}} between substitutions
    and renamings, and then use this resulting \ACode{\ADef{ComposeKit}}
    as an argument for the case with two substitutions.
    As shown below, Agda's instance resolution is up for the task:

    % To get the \ACode{\ADef{ComposeKit}} between substitutions and
    % renamings, the argument \ACode{\ADef{Cᵣ}} is used.
    % The resulting \ACode{\ADef{ComposeKit}} is then again used as an argument
    % for \ACode{\ADef{Cₛ}} to yield the \ACode{\ADef{ComposeKit}}
    % between two substitutions. As shown below, Agda's instance
    % resolution is up for the task:
    \LibCode\KComposeKitInstancesConcrete

@@

  \subsection{Instantiation for System F}
  In this subsection, we show how to instantiate the
  \ACode{\ACon{ComposeTraversal}} abstraction for System F.
  In practice, this is done by our reflection algorithm automatically,
  but it can be instructive to see, as it motivates the axioms of the
  \ACode{\ACon{ComposeKit}}.

  \AppCode\FFusion
  \AppCode\FFusionProofInteresting
  The interesting cases are:
  \begin{itemize}
  \item
    variables, where we need to use the \ACode{\AField{\&/⋯-⋯}} lemma
    provided by the \ACode{\ACon{ComposeKit}}.
  \item
    bindings, where the traversal operation \ACode{\AField{\_⋯\_}}
    needs to lift the map via \ACode{\AField{\_↑\_}}, requiring us to
    distribute the lifting over the composition using
    \ACode{\AField{dist-↑-·}}.
  \end{itemize}

  % \AppCode\FFusionProofRest
  % \AppCode\FComposeTraversal

@@

  We start with \ACode{\ACon{TypingKit}}s, which abstract over
  variable and term \emph{typing}, and then define a
  \ACode{\ACon{TypingTraversal}}, which provides
  substitution-preserves-typing for all \ACode{\ACon{TypingKit}}s.
  \LibCode\KTypingKit
  The operations are completely analogous to those of a \ACode{\ACon{Kit}}, but talk
  about typing instead of terms.
  Building on the fields of a \ACode{\ACon{TypingKit}}, we define further operations:
  \begin{itemize}
  \item
    Using the variable/term typing, we can define a renaming/substitution typing:
    \LibCode\KMapTyping
    \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} states that \ACode{ϕ} is a map that takes
    terms in \ACode{Γ₁} to terms in \ACode{Γ₂}.
  \item 
    Lifting a map preserves its typing:
    \LibCode\KLiftTyping
    % \LibCode\KLiftTypingProof
  \item 
    If a variable/term has a typing, then so does its singleton renaming/substitution:
    \LibCode\KSingleTyping
    % \LibCode\KSingleTypingProof
  \end{itemize}
  To make it easier to explicitly pass typing kits,
  we define the notation \ACode{Γ₂ \ADef{∋*/⊢*[} TK \ADef{]} ϕ \ADef{∶} Γ₁}
  to stand for \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} given some
  \ACode{\ACon{TypingKit}} \ACode{TK}.
  % \LibCode\KTypingNotation

  We then define a \ACode{\ACon{TypingTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  maps, it defines the type preservation of applying maps:
  \LibCode\KTypingTraversal
  Given a term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.
  The large amount of kit-parameters does not impose any restriction,
  as both our \ACode{\ACon{Kit}}s support the \ACode{\ACon{WkKit}}
  extension and can be composed arbitrarily. They only make the previously
  built machinery available to the proof of \ACode{\AField{\_⊢⋯\_}}.

  Given a \ACode{\ACon{TypingTraversal}}, we can easily instantiate the
  \ACode{\ACon{TypingKits}} for variables and terms:
  \LibCode\KTypingInstances
  % \LibCode\KTypingTraversalNotation

  \subsection{Instantiation for System F}
  In this subsection, we show how to instantiate the
  \ACode{\ACon{TypingTraversal}} abstraction for System F.
  This is the only structure that is \emph{not} instantiated
  automatically via reflection, as typing relations can be arbitrary
  complex in general.

  \AppCode\FPreserve
  The interesting parts of the proof are:
  \begin{itemize}
  \item
    There is a strong similarity to the instantiation of
    map traversal \ACode{\AField{\_⋯\_}}. Where \ACode{\AField{\_⋯\_}}
    used \ACode{\AField{`/id}} or \ACode{\AField{\_↑\_}}, we now use their
    preservation lemmas \ACode{\AField{⊢`/id}} or \ACode{\AField{\_∋↑/⊢↑\_}}.
  \item
    The lambda typing constructor \ACode{\ACon{⊢λ}} weakens the type \ACode{t₂} to
    shield it from expression-substitution, requiring us to use the
    \ACode{\AField{⋯-↑-wk}} lemma.
  \item
    The type application constructor \ACode{\ACon{⊢∙}} substitutes \ACode{t₁} into \ACode{t₂},
    requiring us to use the \ACode{\AField{dist-↑-⦅⦆}} lemma.
  \end{itemize}
  Finally, we instantiate the \ACode{\ACon{TypingTraversal}} structure:
  \AppCode\FTypingTraversal

  % As the \ACode{\ACon{Traversal}} and  \ACode{\ACon{ComposeTraversal}} instances
  % can be generally derived via reflection,
  % the \ACode{\AField{\_⊢⋯\_}} lemma is in fact the only lemma that
  % has to been proved manually for subject reduction.
  % From this single lemma follows type preservation for all operations
  % of both renamings and substitutions for all sorts of variables
  % simultaneously.

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism to derive
  instantiations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Syntax}}, \ACode{\ACon{Traversal}} and
  \ACode{\ACon{ComposeTraversal}}.

@@

  \section{Case Studies}
  \label{sec:casestudies}
  Using our full implementation of the framework, we proved subject
  reduction for the following object languages:
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instantiating the
    \ACode{\ACon{TypingTraversal}}.
    Furthermore, the \ACode{\ACon{ComposeKit}} abstraction turned out
    to be useful to prove confluence, allowing to unify certain
    lemmas, which otherwise need to be proved for both renamings and
    substitutions.
    % e.g. that for two terms \ACode{t} and \ACode{t'}
    % with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    % \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    While it would be possible to use our framework only to derive
    untyped substitution and define typing contexts and type preservation
    lemmas by hand, we found an encoding, which still allows us to use the
    the \ACode{\ACon{TypingKit}} abstraction.
    Instead of binding a type variable with a subtyping constraint
    \ACode{α <: t}, we first bind the type variable as \ACode{α : ★}, and
    then bind the constraint as \ACode{c ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization, as constraint variables
    cannot be accesses by the user.
    With that encoding substitution preserves typing is generally true
    again, enabling the use of the framework's typing contexts and
    \ACode{\ACon{TypingKit}} machinery.
    When replacing a type variable \ACode{α <: t} with a type \ACode{t'},
    which is not a subtype of \ACode{t}, then the resulting term is still
    well-typed, but in a context with an unsatisfiable constraint.
  \item
    Object languages with pattern matching can be encoded by adding
    the sorts of the variables bound by a pattern to the pattern sort
    \ACode{\ACon{𝕡}} itself:
    \begin{quote}
    \ACode{\ACon{𝕡} \AKw{:} \ACon{List (Sort Var)} \AKw{→} \ACon{Sort NoVar}} 
    \end{quote}
    This allows the clause of a \ACode{\ACon{match}} expression to
    read the bound variables from the pattern sort and make them
    available as free variables in its body expression.
  \end{itemize}

@@

  \section{Future Work}
  \label{sec:future}
  We want to focus our future work in the following directions:
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Extending the class of typing relations covered by
    the \ACode{\ACon{TypingTraversal}} abstraction and identifying
    a subclass for which it can be derived.
  \item
    Developing a decision procedure for equational
    substitution lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}

@@

  \subsection{Proofs for Section~\ref{sec:composition}}
  \LibCode\KComposeKitAp
  \LibCode\KComposeKitApProof

>>>>

  % Before we define \ACode{\ACon{CKit}} and
  % \ACode{\ACon{CTraversal}}, we need one more ingredient.

@@

  % - Reason about multiple applications in terms of a single application
  % - e.g. wk-cancels-⦅⦆ or dist-↑-⦅⦆, often important for induction hypothesis
  % - Mcbride uses tactics
  % - We introduce an abstraction similar to kits
  % - CKit has primitives for composition, CTraversal has the property

@@

  To fill this gap, we introduce a new abstraction that we call a
  \ACode{\ACon{CKit}},
  which provides an operation \ACode{\AField{\_\&/⋯\_}} that unifies
  \ACode{\ADef{\_\&\_}} and \ACode{\ADef{\_⋯\_}}. This allows us to
  define a general composition, which informally can be described as:
  \ExampleCode\KComposition

  % \ACode{\ACon{CKit}} and \ACode{\ADef{\_·ₖ\_}} are analogous to
  % \ACode{\ACon{Kit}} and \ACode{\ADef{\_→ₖ\_}}, whereas
  % \ACode{\ACon{CTraversal}} and \ACode{\AField{⋯-fusion}} are analogous to
  % \ACode{\ACon{CTraversal}} and \ACode{\AField{\_⋯\_}}.

@@

  In this subsection, we define the \ACode{\ACon{CKit}}
  abstraction, which provides the \ACode{\AField{\_\&/⋯\_}} operation
  that is necessary for defining composition, and the
  \ACode{\ACon{CTraversal}}, which provides the
  \ACode{\AField{⋯-fusion}} lemma for general maps.

  Similarly as it was the case for \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}},
  the lemma of the \ACode{\ACon{CTraversal}} needs to be proved for
  an abstract \ACode{\ACon{CKit}} and in return defines the
  four concrete \ACode{\ACon{CKit}} instances for us. This allows us to break
  the dependencies between the different \ACode{\AField{⋯-fusion}} lemmas,
  similar to how \ACode{\ACon{Kit}} and \ACode{\ACon{Traversal}} allowed
  us to break the dependency between traversals with renamings and substitutions.

  For this purpose, the \ACode{\ACon{CKit}} abstraction, needs
  to satisfy two requirements:
  it needs to provide us with enough machinery to define composition, but
  it also needs to provide us with enough axioms that we are able to instanciate
  the \ACode{\ACon{CTraversal}}. It is defined as followed:
  \LibCode\KCKit
  The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
  (according to \ACode{K₁}) and a renaming or substitution
  (according to \ACode{K₂}) and performs the map traversal returning
  a variable or term (according to \ACode{K₁⊔K₂}).
  The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency and is completely determined by the choice of \ACode{K₁}
  and \ACode{K₂}.

@@

  Similar as for the \ACode{\ACon{Kit}}, we derive further definitions
  from the axioms of a \ACode{\ACon{CKit}}:
  \begin{itemize}
  \item
    We define map composition as motivated by the previous subsection:
    \LibCode\KComposition
  \item
    We show that the \ACode{\AField{\_\&/⋯\_}} operator behaves like
    \ACode{\AField{\_\&\_}} when applied to variables:
    \LibCode\KCKitAp
    % \LibCode\KCKitApProof

    % The \ACode{\AField{`/id}} on the left side is from \ACode{K₁⊔K₂}, whereas
    % the \ACode{\AField{`/id}} on the right side is from \ACode{K₂}.
  \item
    We show that lifting \ACode{\AField{\_↑\_}} distributes over
    composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}
  To make it easier to explicitly specify a \ACode{\ACon{CKit}},
  we define the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}
  to stand for \ACode{ϕ₁ \ADef{·ₖ} ϕ₂} for some
  \ACode{\ACon{CKit}} \ACode{C}.
  % \LibCode\KCKitNotation

  We then define the \ACode{\ACon{CTraversal}} structure,
  which provides the \ACode{\AField{⋯-fusion}} lemma for arbitrary
  \ACode{\ACon{CKit}}s:
  \LibCode\KCTraversal

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the axiom of a \ACode{\ACon{CTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    We define \ACode{\ACon{CKit}} instances for each of the four
    combinations between renamings and substitutions:
    \LibCode\KCKitInstances
    Interestingly, the \ACode{\ADef{Cₛ}} instance has another
    \ACode{\ADef{CKit}} parameter.
    This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
    which yields the \ACode{\ADef{CKit}} between substitutions
    and renamings, and then use this resulting \ACode{\ADef{CKit}}
    as an argument for the case with two substitutions.
    As shown below, Agda's instance resolution is up for the task:

    % To get the \ACode{\ADef{CKit}} between substitutions and
    % renamings, the argument \ACode{\ADef{Cᵣ}} is used.
    % The resulting \ACode{\ADef{CKit}} is then again used as an argument
    % for \ACode{\ADef{Cₛ}} to yield the \ACode{\ADef{CKit}}
    % between two substitutions. As shown below, Agda's instance
    % resolution is up for the task:
    \LibCode\KCKitInstancesConcrete

@@

  \subsection{Instantiation for System F}
  In this subsection, we show how to instantiate the
  \ACode{\ACon{CTraversal}} abstraction for System F.
  In practice, this is done by our reflection algorithm automatically,
  but it can be instructive to see, as it motivates the axioms of the
  \ACode{\ACon{CKit}}.

  \AppCode\FFusion
  \AppCode\FFusionProofInteresting
  The interesting cases are:
  \begin{itemize}
  \item
    variables, where we need to use the \ACode{\AField{\&/⋯-⋯}} lemma
    provided by the \ACode{\ACon{CKit}}.
  \item
    bindings, where the traversal operation \ACode{\AField{\_⋯\_}}
    needs to lift the map via \ACode{\AField{\_↑\_}}, requiring us to
    distribute the lifting over the composition using
    \ACode{\AField{dist-↑-·}}.
  \end{itemize}

  % \AppCode\FFusionProofRest
  % \AppCode\FCTraversal

@@

  We start with \ACode{\ACon{TKit}}s, which abstract over
  variable and term \emph{typing}, and then define a
  \ACode{\ACon{TTraversal}}, which provides
  substitution-preserves-typing for all \ACode{\ACon{TKit}}s.
  \LibCode\KTKit
  The operations are completely analogous to those of a \ACode{\ACon{Kit}}, but talk
  about typing instead of terms.
  Building on the fields of a \ACode{\ACon{TKit}}, we define further operations:
  \begin{itemize}
  \item
    Using the variable/term typing, we can define a renaming/substitution typing:
    \LibCode\KMapTyping
    \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} states that \ACode{ϕ} is a map that takes
    terms in \ACode{Γ₁} to terms in \ACode{Γ₂}.
  \item 
    Lifting a map preserves its typing:
    \LibCode\KLiftTyping
    % \LibCode\KLiftTypingProof
  \item 
    If a variable/term has a typing, then so does its singleton renaming/substitution:
    \LibCode\KSingleTyping
    % \LibCode\KSingleTypingProof
  \end{itemize}
  To make it easier to explicitly pass typing kits,
  we define the notation \ACode{Γ₂ \ADef{∋*/⊢*[} TK \ADef{]} ϕ \ADef{∶} Γ₁}
  to stand for \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} given some
  \ACode{\ACon{TKit}} \ACode{TK}.
  % \LibCode\KTypingNotation

  We then define a \ACode{\ACon{TTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  maps, it defines the type preservation of applying maps:
  \LibCode\KTTraversal
  Given a term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.
  The large amount of kit-parameters does not impose any restriction,
  as both our \ACode{\ACon{Kit}}s support the \ACode{\ACon{WkKit}}
  extension and can be composed arbitrarily. They only make the previously
  built machinery available to the proof of \ACode{\AField{\_⊢⋯\_}}.

  Given a \ACode{\ACon{TTraversal}}, we can easily instantiate the
  \ACode{\ACon{TKits}} for variables and terms:
  \LibCode\KTypingInstances
  % \LibCode\KTTraversalNotation

  \subsection{Instantiation for System F}
  In this subsection, we show how to instantiate the
  \ACode{\ACon{TTraversal}} abstraction for System F.
  This is the only structure that is \emph{not} instantiated
  automatically via reflection, as typing relations can be arbitrary
  complex in general.

  \AppCode\FPreserve
  The interesting parts of the proof are:
  \begin{itemize}
  \item
    There is a strong similarity to the instantiation of
    map traversal \ACode{\AField{\_⋯\_}}. Where \ACode{\AField{\_⋯\_}}
    used \ACode{\AField{`/id}} or \ACode{\AField{\_↑\_}}, we now use their
    preservation lemmas \ACode{\AField{⊢`/id}} or \ACode{\AField{\_∋↑/⊢↑\_}}.
  \item
    The lambda typing constructor \ACode{\ACon{⊢λ}} weakens the type \ACode{t₂} to
    shield it from expression-substitution, requiring us to use the
    \ACode{\AField{⋯-↑-wk}} lemma.
  \item
    The type application constructor \ACode{\ACon{⊢∙}} substitutes \ACode{t₁} into \ACode{t₂},
    requiring us to use \ACode{\AField{dist-↑-⦅⦆}}.
  \end{itemize}
  Finally, we instantiate the \ACode{\ACon{TTraversal}} structure:
  \AppCode\FTTraversal

  % As the \ACode{\ACon{Traversal}} and  \ACode{\ACon{CTraversal}} instances
  % can be generally derived via reflection,
  % the \ACode{\AField{\_⊢⋯\_}} lemma is in fact the only lemma that
  % has to been proved manually for subject reduction.
  % From this single lemma follows type preservation for all operations
  % of both renamings and substitutions for all sorts of variables
  % simultaneously.

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism to derive
  instantiations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Syntax}}, \ACode{\ACon{Traversal}} and
  \ACode{\ACon{CTraversal}}.

@@

  \section{Case Studies}
  \label{sec:casestudies}
  Using our full implementation of the framework, we proved subject
  reduction for the following object languages:
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instantiating the
    \ACode{\ACon{TTraversal}}.
    Furthermore, the \ACode{\ACon{CKit}} abstraction turned out
    to be useful to prove confluence, allowing to unify certain
    lemmas, which otherwise need to be proved for both renamings and
    substitutions.
    % e.g. that for two terms \ACode{t} and \ACode{t'}
    % with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    % \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    While it would be possible to use our framework only to derive
    untyped substitution and define typing contexts and type preservation
    lemmas by hand, we found an encoding, which still allows us to use the
    the \ACode{\ACon{TKit}} abstraction.
    Instead of binding a type variable with a subtyping constraint
    \ACode{α <: t}, we first bind the type variable as \ACode{α : ★}, and
    then bind the constraint as \ACode{c ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization, as constraint variables
    cannot be accesses by the user.
    With that encoding substitution preserves typing is generally true
    again, enabling the use of the framework's typing contexts and
    \ACode{\ACon{TKit}} machinery.
    When replacing a type variable \ACode{α <: t} with a type \ACode{t'},
    which is not a subtype of \ACode{t}, then the resulting term is still
    well-typed, but in a context with an unsatisfiable constraint.
  \item
    Object languages with pattern matching can be encoded by adding
    the sorts of the variables bound by a pattern to the pattern sort
    \ACode{\ACon{𝕡}} itself:
    \begin{quote}
    \ACode{\ACon{𝕡} \AKw{:} \ACon{List (Sort Var)} \AKw{→} \ACon{Sort NoVar}} 
    \end{quote}
    This allows the clause of a \ACode{\ACon{match}} expression to
    read the bound variables from the pattern sort and make them
    available as free variables in its body expression.
  \end{itemize}

@@

  \section{Future Work}
  \label{sec:future}
  We want to focus our future work in the following directions:
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Extending the class of typing relations covered by
    the \ACode{\ACon{TTraversal}} abstraction and identifying
    a subclass for which it can be derived.
  \item
    Developing a decision procedure for equational
    substitution lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}

@@

  \subsection{Proofs for Section~\ref{sec:composition}}
  \LibCode\KCKitAp
  \LibCode\KCKitApProof
